Solar wind prediction for the Solar Probe Plus orbit

### 2017-04-29 paperMVVB/sections/content.tex
created figure 5; adjusted figs. 3,4 and 6; table 4 header; subsections §7 reorganized

		\multicolumn{2}{l}{\multirow{2}{*}{Parameter}}	&\multicolumn{1}{c}{Median}	&\multicolumn{1}{c}{Mean}	&\multicolumn{1}{c}{Balance}	&\multicolumn{1}{c}{MAE}	&\multicolumn{1}{c}{SAR}\\

@@

\subsection{SSN fitting}
\label{sec:ssn_fitting}
To be able to shift the distributions with SSN, add fit a linear SSN dependency to the median
\begin{align}
	x_\text{med}(ssn) &= a_\text{med} \cdot ssn + b_\text{med}\,,	\label{eq:median_with_ssn}
\end{align}
using a factor to the SSN $a_\text{med}$ with a baseline $b_\text{med}$. We relate the mean with a scaling factor to the median to keep the new SSN dependency:
\begin{align}
	x_\text{avg}(ssn) &= (1 + a_\text{avg}) \cdot x_\text{med}(ssn)\,.	\label{eq:mean_with_ssn}
\end{align}
With the implementation of these relations into the lognormal function (\ref{eq:single_lognormal_fit_function}), the new dynamic fit function $W'(x,ssn)$ is then fitted to the yearly data. The three resulting fit coefficients ($a_\text{med}, b_\text{med}$ and $a_\text{avg}$) are presented in Table~\ref{tab:ssn_fit_parameters}.\\
\begin{table*}
	\caption{Resulting fit coefficients from the data fitting with SSN. For the velocity the fit parameters from the double lognormal fit and their balancing function are given. The values in brackets are the estimated standard deviation of each fit parameter.}
	\label{tab:ssn_fit_parameters}
	\centering
	\begin{tabular}{l@{} c@{}
		S[table-format = 1.3(2)e+1]
		S[table-format = 1.4(2)]
		S[table-format = 1.3(2)e+1]
		S[table-format = +1.3(2)e+1]
		c
		}
		\hline\hline
		\multicolumn{2}{l}{\multirow{2}{*}{Parameter}}	&\multicolumn{2}{c}{Median}	&\multicolumn{1}{c}{Mean}	&\multicolumn{2}{c}{Balance}\\
		\cline{3-4}\cline{6-7}
		\multicolumn{2}{l}{}	&\multicolumn{1}{c}{SSN factor $a_\text{med}$\tablefootmark{a}}	&\multicolumn{1}{c}{Baseline $b_\text{med}$\tablefootmark{a}}	&\multicolumn{1}{c}{Scaling factor $a_\text{avg}$}	&\multicolumn{1}{c}{SSN factor $c_a$}	&\multicolumn{1}{c}{Baseline $c_b$}\\
		\hline
		\multicolumn{2}{l}{Magnetic field}	&1.309(19)e-2	&4.285(17)	&8.786(78)e-2	&\multicolumn{1}{c}{--}	&--\\
%		\multicolumn{2}{l}{Velocity}	&2.514(93)e-3	&3.8521(90)	&2.392(26)e-2	&\multicolumn{1}{c}{--}	&--\\
		\multicolumn{2}{l}{Density}	&3.81(25)e-3	&4.495(26)	&3.050(27)e-1	&\multicolumn{1}{c}{--}	&--\\
		\multicolumn{2}{l}{Temperature}	&1.974(26)e-2	&5.729(19)	&6.541(28)e-1	&\multicolumn{1}{c}{--}	&--\\
		\hline
		\multirow{2}{*}{Velocity}	&\multicolumn{1}{c}{$W_1$}	&\multicolumn{1}{c}{--}	&3.633(12)	&1.008(37)e-2	&\multirow{2}{*}{\tablenum{-1.799(95)e-3}}	&\multirow{2}{*}{0.638(32)}\\
			&\multicolumn{1}{c}{$W_2$}	&\multicolumn{1}{c}{--}	&4.831(81)	&2.31(20)e-2	&	&\\
		\hline
	\end{tabular}
	\tablefoot{
		\tablefoottext{a}{Values in their respective units \si{nT}, \SI{e2}{\km\per\s}, \si{\per\cm\cubed} and \SI{e4}{\K}.}
	}
\end{table*}

@@

Again, the velocity gets a special treatment with the double lognormal distribution (\ref{eq:double_lognormal_fit_function}). It is known that slow and fast solar wind stream occurrence rates follow the solar cycle, but their magnitudes stay fairly stable (cite?). This time we keep the two velocity components' positions constant and vary instead their balance with the SSN:
\begin{align}
	c(ssn) &= c_a \cdot ssn + c_b\,.
\end{align}
The fit result (see Table~\ref{tab:ssn_fit_parameters}) is a model in which three years after solar cycle minimum (SSN of zero) the slow solar wind has a share of almost two-thirds and decreases further with increasing SSN (see Fig.~\ref{fig:Vdbl_SSN_ratio_d_plot}).\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/Vdbl_SSN_ratio_d_plot.pdf}}
	\caption{Balance of slow to fast solar wind over the by 3~years lagged SSN. The yearly ratios (+) and their weighted linear fit (solid) are obtained from OMNI data with a threshold of \SI{400}{\km\per\s}. The error bars denote the SSN standard deviation and the relative weight from the yearly data coverage. The model's balance is obtained without a threshold (dashed).}
	\label{fig:Vdbl_SSN_ratio_d_plot}
\end{figure}
% 3 years after minimum (SSN of zero) the slow wind has its maximal share of 0.638(32). 3 years after maximum (SSN of 200) it has only a share of 0.278(37).\\
The ratio of slow to fast wind obtained from yearly OMNI data via simply applying the commonly used constant velocity threshold of \SI{400}{\km\per\s} (cite?) has a flatter SSN dependence than the model's balance. Why?\\
Both should not necessary be identical, because there is no specific velocity threshold between slow and fast solar wind types, as the velocity ranges of both types overlap (cite?).\\

In comparison with the data the fit model roughly matches with the general trend (see Fig.~\ref{fig:OMNI_yearly_BVdblNTSSN_fit_d_plot}).\\

@@

Naturally, we use the double lognormal function~(\ref{eq:double_lognormal_fit_function}) for the velocity distribution fit, resulting in $W''_\text{II}(x,r)$. The additional fit parameters are the balancing parameter $c$ and from the second lognormal part $a_\text{med,2}$ and $a_\text{avg,2}$. The resulting fit coefficients for the four solar wind parameters are presented in Table~\ref{tab:extrapolation_model_fit_parameters}.
\begin{table*}
	\caption{These are the resulting fit coefficients with the single lognormal exponential function, respectively double lognormal for the velocity. The errors in brackets are the estimated standard deviations of each fit parameter.}
	\label{tab:extrapolation_model_fit_parameters}
	\centering
	\sisetup{table-figures-integer=1, table-figures-decimal=3, table-figures-exponent=0}
	\begin{tabular}{l c
	S[table-format = 1.3(2), table-space-text-post = a, table-align-text-post = false]
	S[table-format = 1.3(2), table-space-text-post = a, table-align-text-post = false]
	S[table-format = +1.4(2)]
	S[table-format = 1.3(2)]}
		\hline\hline
		\multicolumn{2}{l}{Parameter}	&\multicolumn{1}{c}{$a_\text{med}$\tablefootmark{a}}	&\multicolumn{1}{c}{$a_\text{avg}$\tablefootmark{a}}	&\multicolumn{1}{c}{$b$}	&$c$\\
		\hline
		\multicolumn{2}{l}{Magnetic field}	&5.358(25)	&5.705(28)	&-1.662(11)	&\multicolumn{1}{c}{--}\\
		\multicolumn{2}{l}{Density}	&5.424(33)	&6.845(47)	&-2.114(20)	&\multicolumn{1}{c}{--}\\
		\multicolumn{2}{l}{Temperature}	&6.357(64)	&10.72(14)	&-1.100(20)	&\multicolumn{1}{c}{--}\\
		\hline
		\multirow{1}{*}{Velocity}	&$W''_1$	&3.707(13)	&3.748(16)	&\multirow{2}{*}{0.0990(51)}	&\multirow{2}{*}{0.557(45)}\\
			&$W''_2$	&5.26(13)	&5.42(11)	&	&\\
		\cline{2-6}
		\multicolumn{2}{r}{$W''_\text{II}$}	&4.35(13)\tablefootmark{b}	&4.77(11)\tablefootmark{b}	&\multicolumn{1}{c}{--}	&\multicolumn{1}{c}{--}\\
		\hline
	\end{tabular}
	\tablefoot{
		\tablefoottext{a}{Values in their respective units \si{nT}, \SI{e2}{\km\per\s}, \si{\per\cm\cubed} and \SI{e4}{\K}.}
		\tablefoottext{b}{Velocity median and mean \SI{1}{\au} values for the resulting function. Error estimates derived from the individual fit part errors.}
	}
\end{table*}

With $c = 0.557$ the velocity balancing parameter is of an expected value similar to that obtained from the Helios time period (the mean SSN during the Helios period was 59, this corresponds to a $c(59) = 0.53$, see Fig.~\ref{fig:Vdbl_SSN_ratio_d_plot}).\\

@@

study time/cycle variations\\
indications for that: fig. + theory + studies/papers...\\
assuming the distance scaling laws are time independent, which we deduced...? which is an approximation.\\

@@

%SSN prediction
choosing SSN prediction sources as input:\\
SIDC prediction---12-month forecast (Kalman fiter combined method)\\
The SWPC prediction (until end of current cycle/end of 2019) follows a consensus of the Solar~Cycle~24 Prediction~Panel.\\
For the prediction of the next solar cycle we simply assume a course similar to the last and thus we simply shift the last cycle by 11~years. Additionally we consider the two alternatives of half and twice the amplitude.\\
SIDC + SWPC + previous cycle~24 (see Fig.~\ref{fig:SPP_orbit_predicted_SSN_overview_c_plot})\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_orbit_predicted_SSN_overview_c_plot.pdf}}
	\caption{SPP's solar distance during its mission time (top). Consecutive Venus flybys bring its perihelia nearer to the Sun. Actual and predicted SSN (bottom), i.e. SIDC 13-month smoothed monthly actual SSN, SIDC prediction, SWPC prediction and simply shifted SSN from previous cycle~24, together with two alternative trends of half and twice the amplitude.}
	\label{fig:SPP_orbit_predicted_SSN_overview_c_plot}
\end{figure}

Implementing the orbital distance data and predicted SSN for the mission time we get SPP's estimated solar wind environment.\\
A zoom into the first and the nearest perihelia shows... (Figs.~\ref{fig:SPP_perihelia_prediction_c_plot} and \ref{fig:SPP_perihelia_prediction_nearest_c_plot}).\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_perihelia_prediction_c_plot.pdf}}
	\caption{Estimated solar wind parameter medians (black) and their error bands (grey) during SPP's first perihelion in 2018. For the velocity the SSN independent slow and fast parts are plotted.}
	\label{fig:SPP_perihelia_prediction_c_plot}
\end{figure}
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_perihelia_prediction_nearest_c_plot.pdf}}
	\caption{Estimated solar wind parameter medians (black) and their error bands (grey) during SPP's nearest perihelion in 2024. For the velocity the SSN independent slow and fast parts are plotted.}
	\label{fig:SPP_perihelia_prediction_nearest_c_plot}
\end{figure}
\subsection{Near-Sun extrapolation behavior/effects}
The extrapolation distance is only about one third of the model range, but as the parameters follow exponential change, one has to look at the logarithmic distance which is indeed one and a half times the model range.\\

@@

\subsection{Model requirements and limits}

>>>>

		\multicolumn{2}{l}{\multirow{2}{*}{Parameter}}	&\multicolumn{1}{c}{Median}	&\multicolumn{1}{c}{Mean}	&\multicolumn{1}{c}{Balance}	&\multicolumn{1}{c}{\multirow{2}{*}{MAE}}	&\multicolumn{1}{c}{SAR}\\

@@

\subsection{SSN fitting}
\label{sec:ssn_fitting}
To be able to shift the distributions with SSN, we add a linear SSN dependency to the median
\begin{align}
	x_\text{med}(ssn) &= a_\text{med} \cdot ssn + b_\text{med}\,,	\label{eq:median_with_ssn}
\end{align}
using a factor to the SSN $a_\text{med}$ with a baseline $b_\text{med}$. We relate the mean with a scaling factor to the median to keep the new SSN dependency:
\begin{align}
	x_\text{avg}(ssn) &= (1 + a_\text{avg}) \cdot x_\text{med}(ssn)\,.	\label{eq:mean_with_ssn}
\end{align}
With the implementation of these relations into the lognormal function (\ref{eq:single_lognormal_fit_function}), the new dynamic fit function $W'(x,ssn)$ is then fitted to the yearly data. The three resulting fit coefficients ($a_\text{med}, b_\text{med}$ and $a_\text{avg}$) are presented in Table~\ref{tab:ssn_fit_parameters}.\\
\begin{table*}
	\caption{Resulting fit coefficients from the data fitting with lagged SSN. For the velocity the fit parameters from the double lognormal fit and their balancing function are given. The values in brackets are the estimated standard deviation of each fit parameter.}
	\label{tab:ssn_fit_parameters}
	\centering
	\begin{tabular}{l@{} c@{}
		S[table-format = 1.3(2)e+1]
		S[table-format = 1.4(2)]
		S[table-format = 1.3(2)e+1]
		S[table-format = +1.3(2)e+1]
		c
		}
		\hline\hline
		\multicolumn{2}{l}{\multirow{2}{*}{Parameter}}	&\multicolumn{2}{c}{Median}	&\multicolumn{1}{c}{Mean}	&\multicolumn{2}{c}{Balance}\\
		\cline{3-4}\cline{6-7}
		\multicolumn{2}{l}{}	&\multicolumn{1}{c}{SSN factor $a_\text{med}$\tablefootmark{a}}	&\multicolumn{1}{c}{Baseline $b_\text{med}$\tablefootmark{a}}	&\multicolumn{1}{c}{Scaling factor $a_\text{avg}$}	&\multicolumn{1}{c}{SSN factor $c_a$}	&\multicolumn{1}{c}{Baseline $c_b$}\\
		\hline
		\multicolumn{2}{l}{Magnetic field}	&1.309(19)e-2	&4.285(17)	&8.786(78)e-2	&\multicolumn{1}{c}{--}	&--\\
%		\multicolumn{2}{l}{Velocity}	&2.514(93)e-3	&3.8521(90)	&2.392(26)e-2	&\multicolumn{1}{c}{--}	&--\\
		\multicolumn{2}{l}{Density}	&3.81(25)e-3	&4.495(26)	&3.050(27)e-1	&\multicolumn{1}{c}{--}	&--\\
		\multicolumn{2}{l}{Temperature}	&1.974(26)e-2	&5.729(19)	&6.541(28)e-1	&\multicolumn{1}{c}{--}	&--\\
		\hline
		\multirow{2}{*}{Velocity}	&\multicolumn{1}{c}{$W_1$}	&\multicolumn{1}{c}{--}	&3.633(12)	&1.008(37)e-2	&\multirow{2}{*}{\tablenum{-1.799(95)e-3}}	&\multirow{2}{*}{0.638(32)}\\
			&\multicolumn{1}{c}{$W_2$}	&\multicolumn{1}{c}{--}	&4.831(81)	&2.31(20)e-2	&	&\\
		\hline
	\end{tabular}
	\tablefoot{
		\tablefoottext{a}{Values in their respective units \si{nT}, \SI{e2}{\km\per\s}, \si{\per\cm\cubed} and \SI{e4}{\K}.}
	}
\end{table*}

@@

The comparison with the yearly data median values shows that the from the model obtained medians have a quite similar slope (see Fig.~\ref{fig:OMNI_yearly_BVNTvsSSN}).\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/OMNI_yearly_BVNTvsSSN.pdf}}
	\caption{Solar wind parameter median over lagged SSN. The yearly data medians (+) with their weighted linear fit (solid) are obtained from OMNI data. The error bars denote the SSN standard deviation and the relative weight from the yearly data coverage. The SSN dependent median (\ref{eq:median_with_ssn}) is from the lognormal model fit (dashed). For the velocity the SSN independent slow and fast model parts are plotted instead (dotted).}
	\label{fig:OMNI_yearly_BVNTvsSSN}
\end{figure}

Again, the velocity gets a special treatment with the double lognormal distribution (\ref{eq:double_lognormal_fit_function}). It is known that slow and fast solar wind stream occurrence rates follow the solar cycle, but their magnitudes stay fairly stable (cite?). This time we keep the two velocity components' positions constant and vary instead their balance with the SSN:
\begin{align}
	c(ssn) &= c_a \cdot ssn + c_b\,.	\label{eq:balance_with_ssn}
\end{align}
The fit result (see Table~\ref{tab:ssn_fit_parameters}) is a model in which three years after solar cycle minimum (SSN of zero) the slow solar wind has a share of almost two-thirds and decreases further with increasing SSN (see Fig.~\ref{fig:Vdbl_SSN_ratio_e_plot}).\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/Vdbl_SSN_ratio_e_plot.pdf}}
	\caption{Balance of slow to fast solar wind over the by 3~years lagged SSN. The yearly ratios (+) and their weighted linear fit (solid) are obtained from OMNI data with a threshold of \SI{400}{\km\per\s}. The error bars denote the SSN standard deviation and the relative weight from the yearly data coverage. The model's balance (\ref{eq:balance_with_ssn}) is obtained without a threshold (dashed).}
	\label{fig:Vdbl_SSN_ratio_e_plot}
\end{figure}
% 3 years after minimum (SSN of zero) the slow wind has its maximal share of 0.638(32). 3 years after maximum (SSN of 200) it has only a share of 0.278(37).\\
The ratio of slow to fast wind obtained from yearly OMNI data via simply applying the commonly used constant velocity threshold of \SI{400}{\km\per\s} (cite?) has a flatter SSN dependence than the model's balance. Why?\\
Both should not necessary be identical, because there is no specific velocity threshold between slow and fast solar wind types, as the velocity ranges of both types overlap (cite?).\\

In comparison with the data the velocity fit model also matches with the general trend (see Fig.~\ref{fig:OMNI_yearly_BVdblNTSSN_fit_d_plot}).\\

@@

Naturally, we use the double lognormal function~(\ref{eq:double_lognormal_fit_function}) for the velocity distribution fit, resulting in $W''_\text{II}(x,r)$. The additional fit parameters are the balancing parameter $c$ and from the second lognormal part $a_\text{med,2}$ and $a_\text{avg,2}$. The resulting fit coefficients for the four solar wind parameters are presented in Table~\ref{tab:extrapolation_model_fit_parameters}.
\begin{table*}
	\caption{These are the resulting fit coefficients with the single lognormal exponential function, respectively double lognormal for the velocity. The errors in brackets are the estimated standard deviations of each fit parameter.}
	\label{tab:extrapolation_model_fit_parameters}
	\centering
	\sisetup{table-figures-integer=1, table-figures-decimal=3, table-figures-exponent=0}
	\begin{tabular}{l c
	S[table-format = 1.3(2), table-space-text-post = a, table-align-text-post = false]
	S[table-format = 1.3(2), table-space-text-post = a, table-align-text-post = false]
	S[table-format = +1.4(2)]
	S[table-format = 1.3(2)]}
		\hline\hline
		\multicolumn{2}{c}{\multirow{2}{*}{Parameter}}	&\multicolumn{1}{c}{Median}	&\multicolumn{1}{c}{Mean}	&\multicolumn{1}{c}{Exponent}	&\multicolumn{1}{c}{Balance}\\
			&	&\multicolumn{1}{c}{$a_\text{med}$\tablefootmark{a}}	&\multicolumn{1}{c}{$a_\text{avg}$\tablefootmark{a}}	&\multicolumn{1}{c}{$b$}	&$c$\\
		\hline
		\multicolumn{2}{l}{Magnetic field}	&5.358(25)	&5.705(28)	&-1.662(11)	&\multicolumn{1}{c}{--}\\
		\multicolumn{2}{l}{Density}	&5.424(33)	&6.845(47)	&-2.114(20)	&\multicolumn{1}{c}{--}\\
		\multicolumn{2}{l}{Temperature}	&6.357(64)	&10.72(14)	&-1.100(20)	&\multicolumn{1}{c}{--}\\
		\hline
		\multirow{1}{*}{Velocity}	&$W''_1$	&3.707(13)	&3.748(16)	&\multirow{2}{*}{0.0990(51)}	&\multirow{2}{*}{0.557(45)}\\
			&$W''_2$	&5.26(13)	&5.42(11)	&	&\\
		\cline{2-6}
		\multicolumn{2}{r}{$W''_\text{II}$}	&4.35(13)\tablefootmark{b}	&4.77(11)\tablefootmark{b}	&\multicolumn{1}{c}{--}	&\multicolumn{1}{c}{--}\\
		\hline
	\end{tabular}
	\tablefoot{
		\tablefoottext{a}{Values in their respective units \si{nT}, \SI{e2}{\km\per\s}, \si{\per\cm\cubed} and \SI{e4}{\K}.}
		\tablefoottext{b}{Velocity median and mean \SI{1}{\au} values for the resulting function. Error estimates derived from the individual fit part errors.}
	}
\end{table*}

With $c = 0.557$ the velocity balancing parameter is of an expected value similar to that obtained from the Helios time period (the mean SSN during the Helios period was 59, this corresponds to a $c(59) = 0.53$, see Fig.~\ref{fig:Vdbl_SSN_ratio_e_plot}).\\

@@

study cycle variation of exponent\\
indications for that: figure! + theory + studies/papers...\\
assuming the distance scaling laws are time independent, which we reason from that; it is an approximation\\

@@

\subsection{SPP mission time frame/SSN prediction}
%SSN prediction
choosing SSN prediction sources as input:\\
SIDC prediction---12-month forecast (Kalman fiter combined method)\\
The SWPC prediction (until end of current cycle/end of 2019) follows a consensus of the Solar~Cycle~24 Prediction~Panel.\\
For the prediction of the next solar cycle we simply assume a course similar to the last and thus we simply shift the last cycle by 11~years. Additionally we consider the two alternatives of half and twice the amplitude.\\
SIDC + SWPC + previous cycle~24 (see Fig.~\ref{fig:SPP_orbit_predicted_SSN_overview_c_plot})\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_orbit_predicted_SSN_overview_c_plot.pdf}}
	\caption{SPP's solar distance during its mission time (top). Consecutive Venus flybys bring its perihelia nearer to the Sun. Actual and predicted SSN (bottom), i.e. SIDC 13-month smoothed monthly actual SSN, SIDC prediction, SWPC prediction and simply shifted SSN from previous cycle~24, together with two alternative trends of half and twice the amplitude.}
	\label{fig:SPP_orbit_predicted_SSN_overview_c_plot}
\end{figure}

=> Implications for the mission...\\

SPP orbit data source\\

\subsection{SPP orbit/near-Sun extrapolation}
The extrapolation distance is only about one third of the model range, but as the parameters follow exponential change, one has to look at the logarithmic distance which is indeed one and a half times the model range.\\

@@

\subsection{SPP solar wind model}
Implementing the orbital distance data and predicted SSN for the mission time we get SPP's estimated solar wind environment.\\
A zoom into the first and the nearest perihelia shows... (Figs.~\ref{fig:SPP_perihelia_prediction_c_plot} and \ref{fig:SPP_perihelia_prediction_nearest_c_plot}).\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_perihelia_prediction_c_plot.pdf}}
	\caption{Estimated solar wind parameter medians (black) and their error bands (grey) during SPP's first perihelion in 2018. For the velocity the SSN independent slow and fast parts are plotted.}
	\label{fig:SPP_perihelia_prediction_c_plot}
\end{figure}
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_perihelia_prediction_nearest_c_plot.pdf}}
	\caption{Estimated solar wind parameter medians (black) and their error bands (grey) during SPP's nearest perihelion in 2024. For the velocity the SSN independent slow and fast parts are plotted.}
	\label{fig:SPP_perihelia_prediction_nearest_c_plot}
\end{figure}
